---
title: "Intro to Time Series Wrangling & Viz"
author: "Kelsie Fronheiser"
date: "2/17/2021"
output: 
  html_document:
    theme: readable
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library (feasts)
library(slider)
library(prettydoc)
```

This data was collected at the Arctic LTER which contains daily obs of air temp
Read in the data:
```{r}
toolik <- read_csv(here("data", "toolikweather.csv"))
```
Convert this into a tsibble
```{r}
#use mdy for month/date/year which is what the data is already in for the column "date"

toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  as_tsibble(key = NULL, index = date)
```

```{r}
ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line()
```

Analysis:
- Seasonality is dominant feature
- No trend
- No major cyclical patterns



Using "index_by()" function to aggregate time series formation by intervals we specify; this is the time series equivalent of "group_by" in dplyr
```{r}

#second row is saying that we want groupings of the entire time series data set (.) by year & month (~yearmonth)
#na.rm = TRUE removes na values

toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% 
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))


ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE))

#this is now one observation per month (averages of daily temps per month)
```

###Aggregate by week
```{r}
toolik_week <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE),
            min_airtemp = min(mean_airtemp))

head(toolik_week)
```


##Filtering Time Series Data
Time series equivalent to filter is 'filter_index()'- use to speify ranges of time series to keep or exclude

```{r}
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")
```

```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")
```

## Season Plots
```{r}
#second line is pulling the year from the date column and keeping obs greater than 2014
toolik_ts %>% 
  filter(year(date) > 2014) %>% 
  gg_season(y = mean_airtemp) +
  theme_light()

toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  theme_light()
```

This helps to capture what is changing (or staying the same) with seasonality across years

## Seasonal subseries plot
```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)
```

The blue lines are the means values


## Find Moving Average
with sliderpackage in R

```{r}
# this creates windows of sections of moving averages 
set.seed(2021)
test <- rnorm(n = 100, mean = 40, sd = 10)


slide(test, ~.x, .before = 2, .after = 2)
```

This will calculate moving average of window size 15
```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp,
                                   mean,
                                   .before = 7,
                                   .after = 7)))

ggplot() +
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp),
            size = 0.2,
            color = "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d)) +
  theme_minimal()
```

## Create the ACF

auto correlation function- visual representation of how obs are correlated with prior obs on the same variable over specified lags of time

```{r}
toolik_ts %>%
  ACF(mean_airtemp) %>% 
  autoplot()

#This shows the days are highly correlated


toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()

# this shows that obs 12 months apart are strongly correlated, 6 or 8 months apart has negative correlation > strong seasonality
#not a lot of dampening happening in terms of auto correlation function 
```


## Decomposition by STL
```{r}

toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

components(toolik_dec) %>% 
  autoplot()
```

The y scale are free - don't be thrown off by shape of panels without considering scale of changes compared to data
You can see here that seasonilty accounts for most residuals , this supports seasonality as most dominant feature 




